---
title: "RenameFiles.Rmd"
author: "Patrick D. lorch"
date: "2024-06-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Rename files with location

We need to make sure file names of .wav files have this format:
LOCATIONorARUID_YYYYMMDD_HHMMSS.WAV

The monitoR package has a function fileCopyRename that does some of this.  We should probably start using this to put sound files into directories on the hard 

This was the .Rmd version that allowed me to develop rnFileWithLoc() and the RunRenameFiles.Rmd.  It may provide a more flexible way to rename files.

```{r rename}
library(stringr)

dir_with_names = "C:/Users/PatrickLorch/OneDrive - SSRS/ARU_bird_callanalysis/BirdNET-Analyzer/testfiles/F1"

# Decide if you want to specify the recorder name or use the enclosing folder
location_name = basename(dir_with_names)
# location_name = "FR1"

rename_file_list = list.files(path = dir_with_names, 
                              pattern = ".wav", 
                              full.names = T)
bn = basename(rename_file_list)
bn1 = paste(location_name, bn, sep = "_")
dn = dirname(rename_file_list)
new_names = file.path(dn, bn1)

file.rename(rename_file_list, new_names)
```

